---
title: "Rmd Template"
subtitle: "Your subtitle here"
author: "Olivia Urbanowski"
date: "`r Sys.Date()`"
urlcolor: blue
output: pdf_document
---


```{r include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,          # don't show code
  warning = FALSE,       # don't show warnings
  message = FALSE,       # don't show messages (less serious warnings)
  cache = FALSE,         # set to TRUE to save results from last compilation
  fig.align = "center",  # center figures
  fig.height=4,          # set default figure size
  fig.width=6
)
library(dplyr)           # load libraries you always use here
library(ggplot2)

set.seed(27)             # make random results reproducible
```

# Synopsis

In a few sentences (max 10), present what this report is about.

# Problem 1

Here is a [link](https://adv-r.hadley.nz/). Your homework goes here.

Downloading a file: 
```{r}
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
if(!file.exists("HAR Dataset.zip")) {
    download.file(fileUrl,"HAR Dataset.zip", mode="wb" )
}
```

Equations and special characters are nicely presented this way: $\mu= 1/\lambda$  
For inline code, either `hello world` shows the code but doesn't execute it, or `r sys.Date()` will print the date. 

Also, some code:

```{r}
## here goes your code
```

Yet more text, and another code section:

```{r}
knitr::kable(head(select(au, 1:5)),  caption = "Sample vs. Theory")
```
\pagebreak

# Code

```{r ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
# this R markdown chunk generates a code appendix
```