---
title: "Internet use in Europe"
subtitle: "Coursera course project - Developing data products"
author: "ourbanow"
date: "6/15/2020"
output:
  ioslides_presentation: default
---

## Introduction

The European Union collects since 2002 data from countries around Europe (EU-members and some non-EU-members) about the relationship of citizens with the internet.  
Datasets are free to access and use at this address:  
[European Commission website](https://ec.europa.eu/info/statistics_en)  
This data, albeit being of unequal quality depending on the country, the date and the type of collected data, is a great source of information for whoever would like to get some insights about the evolution of the use of internet in Europe over the last 2 decades.  
With this in mind, we offer you acces to a [shinyapp](https://ourbanow.shinyapps.io/EUInternetuse/) which will help you visualize and analyze the evolution of the use of internet per type of activity, per age, per country.  

## Example

```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(dplyr)          
library(ggplot2)
library(plotly)
activities <- read.csv(url("https://raw.githubusercontent.com/ourbanow/DataScienceCoursera/master/Developing%20Data%20Products/Internetuse.csv"))
names(activities)[1] <- "ExpPeriod"
activity <- "sending/receiving e-mails"
age <- "all ages"
countrylist <- c("European Union", "Germany", "France", "Poland", 
                 "Spain", "United Kingdom", "Italy")
mydata <- activities %>% 
    filter(ExpCountry %in% countrylist) %>% 
    filter(ExpBrkDwn == age) %>%
    filter(Internetuse == activity)
mytitle <- paste("Internet use:", activity, "for", age)
mysubtitle <- "In a few selected countries"
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
g <- ggplot(mydata, aes(x= ExpPeriod, y = Value, colour=
    ExpCountry)) + geom_line(    ) + scale_x_continuous(
    name="Year", breaks = seq(2002,2020,2), limits=c(2002,
    2019)) + scale_y_continuous(name="Percentage", breaks =
    seq(0, 100,10), limits=c(0, 100)) + ggtitle(mytitle, 
    subtitle =mysubtitle) + theme(plot.title = element_text(
    hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)
    ) + labs(colour="Country")
ggplotly(g)
```

## Code 1/2

```{r ref.label = knitr::all_labels(results == 'hide'), echo = TRUE, eval = FALSE}
```

## Code 2/2

```{r ref.label = knitr::all_labels(results != 'hide'), echo = TRUE, eval = FALSE}
```