---
title: "Practical Machine Learning course project"
subtitle: "Predicting the correctness of barbell lifts execution"
author: "ourbanow"
date: "6/10/2020"
output: 
  html_document: 
    keep_md: yes
---

```{r include = FALSE}
## This chunk allows all the code to be sent to the appendix
knitr::opts_chunk$set( echo = FALSE, warning = FALSE,  message = FALSE,  cache = TRUE )
library(dplyr)          
library(ggplot2)
library(caret)
library(rattle)

set.seed(23)            
```

## Executive summary

The purpose of this document is to try and predict the correctness of the execution of barbell lift exercises.  
The initial study is available here:  [Qualitative Activity Recognition of Weight Lifting Exercises](http://groupware.les.inf.puc-rio.br/public/papers/2013.Velloso.QAR-WLE.pdf).  

For this study, we used the same [initial dataset](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv), but we used our own approach for building a prediction algorithm.  
The R code can be found in the appendix.  
The results run on the test set can be found in the Final Validation section.

## Data preparation 
To perform the analysis, six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), or with an "error" (Classes B through E). Their physical activity was recorded using different sensors; the results are available in a [dataset](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv) under Creative Common license (CC BY-SA).  

```{r}
 ## downloading and loading the dataset
fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
if(!file.exists("pml-training.csv")) {
    download.file(fileUrl,"pml-training.csv", mode="wb" )
}
trainset <- read.csv("pml-training.csv")
```

This dataset consists of 19 622 observations of 160 variables. Unfortunately, many observations are uncomplete - only `r sum(complete.cases(trainset))` are actually complete. We also do not need all the variables for the predicitions, so the first step is to clean the dataset.

```{r}
## cleaning dataset
training <- trainset %>%
  select(roll_belt:classe) %>% ## removing the names of the participants and the time stamps
    select(-c(kurtosis_roll_belt:var_yaw_belt)) %>% ## removing variables which have only 1 value per window
      select(-c(var_accel_arm:var_yaw_arm)) %>%
        select(-c(kurtosis_roll_arm:amplitude_yaw_arm)) %>%
          select(-c(kurtosis_roll_dumbbell:amplitude_yaw_dumbbell)) %>%
            select(-c(var_accel_dumbbell:var_yaw_dumbbell)) %>%
              select(-c(kurtosis_roll_forearm:amplitude_yaw_forearm)) %>% 
                select(-c(var_accel_forearm:var_yaw_forearm))
```
 
Once cleaned, we are left with `r dim(training)[2]` variables, which is a lot. A further analysis of correlated variables might determine how many of them are correlated, using a 90% threshold:    

```{r}
## checking correlation of left variables
corr <- abs(cor(training[-53]))
(sum(corr >= 0.90)-52)/2
```

We could remove them. However, the methods used to build a prediction model will be partition algorithms; therefore, it does not appear compulsory to reduce the number of variables.  

## Model building
We built 3 different models, trained on a subset of the training set, and validated on a "validation" subset of the training set.  

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.75, list=FALSE)
subtrain <- training[inTrain,]
subvalid <- training[-inTrain,]
```

### 1. Recursive partitioning
The first basic model used is recursive partitioning.  
```{r}
start.time1 <- Sys.time()
rpartmod <- train(classe ~., data = subtrain, method = "rpart")
end.time1 <- Sys.time()
time.taken1 <- end.time1 - start.time1
```

Time taken to execute this model: 
```{r}
time.taken1
```

Applying this model to our validation set produces the following results:  
```{r}
rpartpredict <- predict(rpartmod, newdata=subvalid)
Crpart <- confusionMatrix(subvalid$classe, rpartpredict)
Crpart$overall[1]
grpart <- ggplot(data = data.frame(Crpart$table))
grpart + geom_col(aes(x=Reference, y=Freq, fill=Prediction)) + labs(title = "Validation results with the rpart method") + theme(plot.title = element_text(hjust = 0.5))
```

### 2. Bagging
The second model used is bootstrap aggregating. Time taken to execute this model:    
```{r}
start.time2 <- Sys.time()
baggmod <- train(classe ~., data = subtrain, method = "treebag")
end.time2 <- Sys.time()
time.taken2 <- end.time2 - start.time2
time.taken2
```

Applying this model to our validation set produces the following results:  
```{r}
baggpredict <- predict(baggmod, newdata=subvalid)
Cbagg <- confusionMatrix(subvalid$classe, baggpredict)
Cbagg$overall[1]
gbagg <- ggplot(data = data.frame(Cbagg$table))
gbagg + geom_col(aes(x=Reference, y=Freq, fill=Prediction)) + labs(title = "Validation results with the bagging method") + theme(plot.title = element_text(hjust = 0.5))
```

### 3. Random forest
The third model used is a random forest.  
```{r}
rfControl <- trainControl(method = "cv", number = 3, allowParallel = TRUE, verboseIter=FALSE) ## reducing runtime
start.time3 <- Sys.time()
rfmod <- train(classe ~., data = subtrain, method = "rf", trControl = rfControl)
end.time3 <- Sys.time()
time.taken3 <- end.time3 - start.time3
```

Time taken to execute this model: 
```{r}
time.taken3
```

Applying this model to our validation set produces the following results:  
```{r}
rfpredict <- predict(rfmod, newdata=subvalid)
Crf <- confusionMatrix(subvalid$classe, rfpredict)
Crf$overall[1]
grf <- ggplot(data = data.frame(Crf$table))
grf + geom_col(aes(x=Reference, y=Freq, fill=Prediction)) + labs(title = "Validation results with the rf method") + theme(plot.title = element_text(hjust = 0.5))
```

### 4. General Boosting model
The fourth model used is gbm.  
```{r, results='hide'}
gbmControl <- trainControl(method = "repeatedcv", number = 5, repeats = 1) ## reducing runtime
start.time4 <- Sys.time()
gbmmod <- train(classe ~., data = subtrain, method = "gbm", trControl = gbmControl)
end.time4 <- Sys.time()
time.taken4 <- end.time4 - start.time4
```

Time taken to execute this model: 
```{r}
time.taken4
```

Applying this model to our validation set produces the following results:  
```{r}
gbmpredict <- predict(gbmmod, newdata=subvalid)
Cgbm <- confusionMatrix(subvalid$classe, gbmpredict)
Cgbm$overall[1]
ggbm <- ggplot(data = data.frame(Cgbm$table))
ggbm + geom_col(aes(x=Reference, y=Freq, fill=Prediction)) + labs(title = "Validation results with the GBM method") + theme(plot.title = element_text(hjust = 0.5))
```

## Model comparison
On our validation dataset, the four methods perfom as following:
```{r}
methods <- c("RPART", "BAGGING", "RF", "GBM")
runtimes <- c(round(time.taken1, 2), round(time.taken2, 2), round(time.taken3, 2), round(time.taken4, 2))
accuracy <- c(round(Crpart$overall[1],2), round(Cbagg$overall[1],2), round(Crf$overall[1],2), round(Cgbm$overall[1],2))
compar <- rbind(runtimes, accuracy)
colnames(compar) <- methods
rownames(compar) <- c("Runtime (s)", "Accuracy")
compar
```

The basic recursive partitioning method is fast, but has a quite low accuracy. All three other methods have very high accuracy, but long runtimes.  As a compromise, we select the Bagging method to be applied on the test dataset.

## Final validation
Applying the developped Bagging model on a test dataset provides the following results:  
```{r}
## downloading and loading the dataset
fileUrl2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
if(!file.exists("pml-testing.csv.csv")) {
    download.file(fileUrl2,"pml-testing.csv", mode="wb" )
}
testset <- read.csv("pml-testing.csv")
```
```{r}
testpredict <- predict(baggmod, newdata=testset)
testpredict
```

## Code

```{r ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}

```